---
title: "Crop Yield Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Crop Yield Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/patelris/crop-yield-prediction-dataset\>*

### Reference:

*\<Food and Agriculture Organization & World Data Bank. (n.d.). Crop Yield Prediction Dataset. Retrieved from http://www.fao.org/home/en/ and https://data.worldbank.org/\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r EDA}
# Load dataset
crop_data <- read.csv("yield_df.csv", colClasses = c(
  id = "integer",
  Area = "factor",
  Item = "factor",
  Year = "integer",
  area_yield = "integer",
  average_rain_fall_mm_per_year = "integer",
  pesticides_tonnes = "numeric",
  avg_temp = "numeric"
))

# Display the structure of the dataset
str(crop_data)

# View the first few rows of the dataset
head(crop_data)

# View the dataset in a separate viewer window
View(crop_data)
```

## Measures of Frequency
```{r MOF}
# Calculate frequency of each category in 'Area' column
area_freq <- table(crop_data$Area)
print(area_freq)

# Calculate frequency of each category in 'Item' column
item_freq <- table(crop_data$Item)
print(item_freq)

# Summary statistics for 'area_yield' column
summary(crop_data$area_yield)

# Summary statistics for 'average_rain_fall_mm_per_year' column
summary(crop_data$average_rain_fall_mm_per_year)

# Summary statistics for 'pesticides_tonnes' column
summary(crop_data$pesticides_tonnes)

# Summary statistics for 'avg_temp' column
summary(crop_data$avg_temp)
```

## Measures of Central Tendency
```{r MOCT}
# Define a function to calculate mode
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Calculate mean, median, and mode for numerical variables

# Mean, median, and mode for 'area_yield' column
mean_area_yield <- mean(crop_data$area_yield)
median_area_yield <- median(crop_data$area_yield)
mode_area_yield <- Mode(crop_data$area_yield)

# Mean, median, and mode for 'average_rain_fall_mm_per_year' column
mean_rainfall <- mean(crop_data$average_rain_fall_mm_per_year)
median_rainfall <- median(crop_data$average_rain_fall_mm_per_year)
mode_rainfall <- Mode(crop_data$average_rain_fall_mm_per_year)

# Mean, median, and mode for 'pesticides_tonnes' column
mean_pesticides <- mean(crop_data$pesticides_tonnes)
median_pesticides <- median(crop_data$pesticides_tonnes)
mode_pesticides <- Mode(crop_data$pesticides_tonnes)

# Mean, median, and mode for 'avg_temp' column
mean_temp <- mean(crop_data$avg_temp)
median_temp <- median(crop_data$avg_temp)
mode_temp <- Mode(crop_data$avg_temp)

# Print results
cat("Mean, Median, and Mode for 'area_yield' column:\n")
cat("Mean:", mean_area_yield, "Median:", median_area_yield, "Mode:", mode_area_yield, "\n\n")

cat("Mean, Median, and Mode for 'average_rain_fall_mm_per_year' column:\n")
cat("Mean:", mean_rainfall, "Median:", median_rainfall, "Mode:", mode_rainfall, "\n\n")

cat("Mean, Median, and Mode for 'pesticides_tonnes' column:\n")
cat("Mean:", mean_pesticides, "Median:", median_pesticides, "Mode:", mode_pesticides, "\n\n")

cat("Mean, Median, and Mode for 'avg_temp' column:\n")
cat("Mean:", mean_temp, "Median:", median_temp, "Mode:", mode_temp, "\n\n")
```

## Measures of Distribution
```{r MOD}
# Load required library for skewness and kurtosis calculation
library(e1071)

# Calculate measures of distribution for numerical variables

# Variance and standard deviation for 'area_yield' column
var_area_yield <- var(crop_data$area_yield)
sd_area_yield <- sd(crop_data$area_yield)

# Variance and standard deviation for 'average_rain_fall_mm_per_year' column
var_rainfall <- var(crop_data$average_rain_fall_mm_per_year)
sd_rainfall <- sd(crop_data$average_rain_fall_mm_per_year)

# Variance and standard deviation for 'pesticides_tonnes' column
var_pesticides <- var(crop_data$pesticides_tonnes)
sd_pesticides <- sd(crop_data$pesticides_tonnes)

# Variance and standard deviation for 'avg_temp' column
var_temp <- var(crop_data$avg_temp)
sd_temp <- sd(crop_data$avg_temp)

# Skewness and kurtosis for 'area_yield' column
skew_area_yield <- skewness(crop_data$area_yield)
kurt_area_yield <- kurtosis(crop_data$area_yield)

# Skewness and kurtosis for 'average_rain_fall_mm_per_year' column
skew_rainfall <- skewness(crop_data$average_rain_fall_mm_per_year)
kurt_rainfall <- kurtosis(crop_data$average_rain_fall_mm_per_year)

# Skewness and kurtosis for 'pesticides_tonnes' column
skew_pesticides <- skewness(crop_data$pesticides_tonnes)
kurt_pesticides <- kurtosis(crop_data$pesticides_tonnes)

# Skewness and kurtosis for 'avg_temp' column
skew_temp <- skewness(crop_data$avg_temp)
kurt_temp <- kurtosis(crop_data$avg_temp)

# Print results
cat("Measures of Distribution for 'area_yield' column:\n")
cat("Variance:", var_area_yield, "Standard Deviation:", sd_area_yield, "Skewness:", skew_area_yield, "Kurtosis:", kurt_area_yield, "\n\n")

cat("Measures of Distribution for 'average_rain_fall_mm_per_year' column:\n")
cat("Variance:", var_rainfall, "Standard Deviation:", sd_rainfall, "Skewness:", skew_rainfall, "Kurtosis:", kurt_rainfall, "\n\n")

cat("Measures of Distribution for 'pesticides_tonnes' column:\n")
cat("Variance:", var_pesticides, "Standard Deviation:", sd_pesticides, "Skewness:", skew_pesticides, "Kurtosis:", kurt_pesticides, "\n\n")

cat("Measures of Distribution for 'avg_temp' column:\n")
cat("Variance:", var_temp, "Standard Deviation:", sd_temp, "Skewness:", skew_temp, "Kurtosis:", kurt_temp, "\n\n")
```

## Measures of relationship
```{r MOR}
# Calculate Pearson correlation coefficients between numerical variables
correlation_matrix <- cor(crop_data[, c("area_yield", "average_rain_fall_mm_per_year", "pesticides_tonnes", "avg_temp")])
print(correlation_matrix)

# Create contingency table for categorical variables 'Area' and 'Item'
contingency_table <- table(crop_data$Area, crop_data$Item)
print(contingency_table)

# Perform chi-square test of independence
chi_square_test <- chisq.test(contingency_table)
print(chi_square_test)
```

## ANOVA
```{r ANOVA}
# Load required library for ANOVA
library(stats)

# Perform ANOVA for numerical variable 'area_yield' across different levels of 'Item' (crop type)
anova_result <- aov(area_yield ~ Item, data = crop_data)
print(summary(anova_result))
```

## Plots
```{r Plots}
# Load required libraries for plotting
library(ggplot2)
library(gridExtra)

# Univariate plots for numerical variables

# Histogram for 'area_yield'
hist_area_yield <- ggplot(crop_data, aes(x = area_yield)) +
  geom_histogram(binwidth = 1000, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Area Yield",
       x = "Area Yield",
       y = "Frequency")

# Histogram for 'average_rain_fall_mm_per_year'
hist_rainfall <- ggplot(crop_data, aes(x = average_rain_fall_mm_per_year)) +
  geom_histogram(binwidth = 50, fill = "lightgreen", color = "black") +
  labs(title = "Histogram of Average Rainfall",
       x = "Average Rainfall (mm/year)",
       y = "Frequency")

# Histogram for 'pesticides_tonnes'
hist_pesticides <- ggplot(crop_data, aes(x = pesticides_tonnes)) +
  geom_histogram(binwidth = 10, fill = "lightcoral", color = "black") +
  labs(title = "Histogram of Pesticides Usage",
       x = "Pesticides (tonnes)",
       y = "Frequency")

# Histogram for 'avg_temp'
hist_temp <- ggplot(crop_data, aes(x = avg_temp)) +
  geom_histogram(binwidth = 1, fill = "lightyellow", color = "black") +
  labs(title = "Histogram of Average Temperature",
       x = "Average Temperature (°C)",
       y = "Frequency")

# Combine histograms into a single plot grid
histogram_grid <- grid.arrange(hist_area_yield, hist_rainfall, hist_pesticides, hist_temp, ncol = 2)

# Univariate plots for categorical variables

# Bar plot for 'Area'
bar_area <- ggplot(crop_data, aes(x = Area)) +
  geom_bar(fill = "lightblue") +
  labs(title = "Bar Plot of Areas",
       x = "Area",
       y = "Count")

# Bar plot for 'Item'
bar_item <- ggplot(crop_data, aes(x = Item)) +
  geom_bar(fill = "lightgreen") +
  labs(title = "Bar Plot of Items",
       x = "Item",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better visibility

# Combine bar plots into a single plot grid
barplot_grid <- grid.arrange(bar_area, bar_item, ncol = 2)
```

# Preprocessing and Data Transformation
## Missing values
```{r }
# Remove the first column from the dataset
crop_data <- crop_data[, -1]

# Display the first few rows of the modified dataset
head(crop_data)


# Check for missing values in the dataset
missing_values <- any(is.na(crop_data))
print(missing_values)
```

## Transformation
```{r}
# Load required library for data transformation
library(dplyr)

# Check the distribution of numerical variables before transformation
summary(crop_data[, c("area_yield", "average_rain_fall_mm_per_year", "pesticides_tonnes", "avg_temp")])

# Perform log transformation on skewed numerical variables
crop_data_transformed <- crop_data %>%
  mutate(
    area_yield_log = log(area_yield + 1),  # Adding 1 to handle zero values
    average_rain_fall_mm_per_year_log = log(average_rain_fall_mm_per_year + 1),
    pesticides_tonnes_log = log(pesticides_tonnes + 1),
    avg_temp_log = log(avg_temp + 1)
  )

# Check the distribution of transformed numerical variables
summary(crop_data_transformed[, c("area_yield_log", "average_rain_fall_mm_per_year_log", "pesticides_tonnes_log", "avg_temp_log")])

```

# Training Model
## Data Splitting
```{r}
# Load required library for data splitting
library(caret)

# Set seed for reproducibility
set.seed(123)

# Define the proportion of data to be used for training (e.g., 80%)
train_proportion <- 0.8

# Perform data splitting
train_index <- createDataPartition(crop_data_transformed$area_yield, p = train_proportion, list = FALSE)
train_data <- crop_data_transformed[train_index, ]
test_data <- crop_data_transformed[-train_index, ]

# Check the dimensions of the training and testing sets
cat("Training data dimensions:", dim(train_data), "\n")
cat("Testing data dimensions:", dim(test_data), "\n")
```

## Bootstrapping
```{r}
# Load required library for bootstrapping
library(boot)

# Define the function to calculate the statistic of interest (e.g., mean)
# For example, let's calculate the mean of 'area_yield'
boot_mean <- function(data, indices) {
  mean(data[indices, "area_yield"])
}

# Set the number of bootstrap samples
num_bootstrap_samples <- 1000

# Perform bootstrapping
boot_results <- boot(crop_data_transformed, boot_mean, R = num_bootstrap_samples)

# Display the results
print(boot_results)
```

## Cross-validation
```{r}
# Load required libraries for cross-validation
library(caret)

# Define the training control parameters
train_control <- trainControl(method = "cv",   # "cv" for k-fold cross-validation
                              number = 10)     # Number of folds (e.g., 10-fold cross-validation)

# Define the model training process (e.g., linear regression)
model <- train(area_yield ~ .,                     # Formula for the model
               data = crop_data_transformed,       # Dataset
               method = "lm",                      # Method (e.g., linear regression)
               trControl = train_control)          # Training control parameters

# Print the cross-validation results
print(model)
```

## Training Model
```{r}
# Load required libraries for modeling
library(caret)

# Train linear regression model
lm_model <- train(area_yield ~ .,                     # Formula for the model
                  data = crop_data_transformed,       # Dataset
                  method = "lm",                      # Method (linear regression)
                  trControl = trainControl(method = "cv", number = 10))  # Training control parameters

# Print the linear regression model results
print(lm_model)

# Train decision tree model
rpart_model <- train(area_yield ~ .,                     # Formula for the model
                     data = crop_data_transformed,       # Dataset
                     method = "rpart",                   # Method (decision trees)
                     trControl = trainControl(method = "cv", number = 10))  # Training control parameters

# Print the decision tree model results
print(rpart_model)

# Train GBM model
gbm_model <- train(area_yield ~ .,                     # Formula for the model
                   data = crop_data_transformed,       # Dataset
                   method = "gbm",                    # Method (gradient boosting machines)
                   trControl = trainControl(method = "cv", number = 10))  # Training control parameters

# Print the GBM model results
print(gbm_model)
```

## Performance comparison
```{r}
# Load required libraries for modeling
library(caret)

# Define the training control parameters
train_control <- trainControl(method = "cv",   # "cv" for k-fold cross-validation
                              number = 10,     # Number of folds for cross-validation
                              verboseIter = TRUE)  # Print progress during model training

# Define the models to train
models <- c("lm", "rpart", "gbm")

# Train the models
results <- lapply(models, function(model) {
  train(area_yield ~ .,                     # Formula for the model
        data = crop_data_transformed,       # Dataset
        method = model,                     # Method (e.g., linear regression, decision trees, etc.)
        trControl = train_control)          # Training control parameters
})

# Compare model performance using resamples
model_resamples <- resamples(results)

# Summarize model performance
summary(model_resamples)
```

## Saving Model
```{r}
# Load the saved model
loaded_gbm_model <- readRDS("./models/saved_gbm_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  Area = "Albania",
  Item = "Maize",
  Year = 1990,
  average_rain_fall_mm_per_year = 1485,
  pesticides_tonnes = 121,
  avg_temp = 16.37,
  area_yield_log = log(36613 + 1),  # Apply the same transformation as during model training
  average_rain_fall_mm_per_year_log = log(1485 + 1),  # Apply the same transformation as during model training
  pesticides_tonnes_log = log(121 + 1),  # Apply the same transformation as during model training
  avg_temp_log = log(16.37 + 1)  # Apply the same transformation as during model training
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_gbm_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

